### 2022-12-05 chapter1/chapter1.tex
Updated alignment trimming methods and fixed inconsistency in number style

However, connected components and triangle clustering are special cases of the more general \textit{k}-clique percolation algorithm where \textit{k} equals two and three, respectively. The clique percolation algorithm detects clusters by first identifying cliques, sets of nodes which are fully connected, of a specified size \textit{k} in the graph (Fig.~\ref{fig:percolation}A). Clusters are then taken as the connected components of an overlap graph where an edge exists between two cliques if they share \textit{k}-1 nodes in common. An intuitive way to visualize this algorithm is by ``rolling'' a clique of some size \textit{k} over the graph (Fig.~\ref{fig:percolation}B). More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified. The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique. A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster. The strength of this algorithm is its ability to exclude sparsely connected nodes from clusters with an easily tunable parameter \textit{k}. Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships (Fig.~\ref{fig:percolation}C). We set \textit{k} to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).

@@

\subsection*{Alignment refinement}
The representative sequences in the single copy orthologous groups were aligned with the same settings as described in the previous section except \textit{a\textsubscript{max}} was set to 0.7. A binary profile was created from the alignment using Gaussian process sequence weighting as described in the previous section. (Because the orthologous groups were single copy and contained only representative sequences, each sequence received the full weight associated with its species.) The binary profile was then converted into a binary mask by identifying where the weighted fraction of non-gap symbols exceeded 0.5. The binary mask was closed with a structuring element of size three, and highly conserved regions were identified as the contiguous intervals of this closed mask with a minimum length of 10. Diverged regions were taken as the complement of the highly conserved regions. For each diverged region, the corresponding segments of the sequences in the initial alignment were extracted and aligned with \textit{a\textsubscript{max}} set to 0.4. The resulting sub-alignments were then stitched into the initial alignment.

@@

The posterior distributions over states were computed for both the insertion and missing data phylo-HMMs. To trim the alignments, the posterior distributions from the insertion phylo-HMM were converted to state assignments with the following method. First the probabilities for states 2 and 3 were combined. Since this probability can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries as the columns where the gap pattern changes most abruptly. Instead, a high cutoff of 0.9 was used to define a ``seed'' region. The boundaries of the seed were then expanded both inwards and outwards to define intervals from which to select boundaries. The outward expansion halted when the probability or its derivative was below 0.05 and 0.001, respectively. The inward expansion halted when the derivative was below 0.02. The left and right boundaries were chosen as the columns in each expansion with the maximum product between the derivative and the number of sequences which changed to or from a gap between columns. By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries. In contrast, the posterior probabilities from the missing data phylo-HMM changed sharply, so positions were labeled as the ``missing data'' state if the posterior probability exceeded 0.95. These assignments can be used to filter segments or entire sequences from downstream analyses and are available in the supplementary data.

>>>>

However, connected components and triangle clustering are special cases of the more general \textit{k}-clique percolation algorithm where \textit{k} equals two and three, respectively. The clique percolation algorithm detects clusters by first identifying cliques, sets of nodes which are fully connected, of a specified size \textit{k} in the graph (Fig.~\ref{fig:percolation}A). Clusters are then taken as the connected components of an overlap graph where an edge exists between two cliques if they share \textit{k}-1 nodes in common. An intuitive way to visualize this algorithm is by ``rolling'' a clique of some size \textit{k} over the graph (Fig.~\ref{fig:percolation}B). More specifically, a cluster is initiated when a set of nodes which form a \textit{k}-clique is identified. The cluster expands by shifting the \textit{k}-clique to an adjacent \textit{k}-clique that shares \textit{k}-1 nodes in common with the current \textit{k}-clique. A cluster stops expanding when there are no adjacent \textit{k}-cliques, and the algorithm terminates when there are no \textit{k}-cliques which are not part of a cluster. The strength of this algorithm is its ability to exclude sparsely connected nodes from clusters with an easily tunable parameter \textit{k}. Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships (Fig.~\ref{fig:percolation}C). We set \textit{k} to four as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).

@@

\subsection*{Alignment refinement}
The representative sequences in the single copy orthologous groups were aligned with the same settings as described in the previous section except \textit{a\textsubscript{max}} was set to 0.7. A binary profile was created from the alignment using Gaussian process sequence weighting as described in the section on selecting representative sequences. (Because the orthologous groups were single copy and contained only representative sequences, each sequence received the full weight associated with its species.) The binary profile was then converted into a binary mask by identifying where the weighted fraction of non-gap symbols exceeded 0.5. The binary mask was closed with a structuring element of size three, and highly conserved regions were identified as the contiguous intervals of this closed mask with a minimum length of 10. Diverged regions were taken as the complement of the highly conserved regions. For each diverged region, the corresponding segments of the sequences in the initial alignment were extracted and aligned with \textit{a\textsubscript{max}} set to 0.4. The resulting sub-alignments were then stitched into the initial alignment.

@@

The posterior distributions over states were calculated for each alignment using the trained insertion phylo-HMM. States 2 and 3 have distinct characteristics which require different trimming strategies, so regions with a high probability of state 3 were handled first. Because the posterior probability of a state can change rapidly or gradually depending on the local context, a simple cutoff would not necessarily define the boundaries as the columns where the gap pattern changes most abruptly. Instead, a high cutoff of 0.95 was used to define a ``seed'' region. The left and right endpoints of the seed were then expanded both inwards and outwards to define two intervals from which boundaries were selected. The outward expansion halted when the probability or its derivative was below 0.01 and 0.001, respectively. The inward expansion halted when the derivative was below 0.001. The left and right boundaries were chosen as the columns in each interval with the maximum product between the derivative and the change in the gap profile between columns. The gap profile was calculated as the number of gaps in each column using Gaussian process sequence weighting as described in the section on selecting representative sequences. By combining where the model’s confidence changed rapidly with the observed change in the gap pattern, this method generally selected reasonable boundaries.

Because the long poorly supported segments in state 3 regions were sometimes aligned to highly conserved columns or short segments in other sequences, trimming columns entirely would remove these segments from the other sequences even if they do not qualify as long and poorly supported themselves. Thus, regions with high state 3 probabilities were trimmed at the level of individual sequences rather than entire columns with the following method. First, the mean number of non-gap symbols in a region was calculated using Gaussian process sequence weighting. (The sequences with the five most non-gap symbols were also excluded to not bias the estimate with long poorly supported segments.) The final mean $\mu$ was taken as the minimum of this value and two. A cutoff \textit{k}, derived from a geometric model of the number of non-gap symbols and a significance level $\alpha$, was calculated using the following equation $k = \log(\alpha) / \log(1 - p) - 1$ where $p = \frac{1}{\mu + 1}$ and $\alpha = 0.01$. Any sequence whose number of non-gap symbols in the region equaled or exceeded this value was trimmed by replacing all non-gap symbols with gaps.

To trim the remaining state 2 regions, the posterior probability of state 2 was added to a modified state 3 probability which was set to zero for any state 3 regions identified in the previous step. This ensured that any regions which were intermediate between state 2 and 3 were included. Regions were defined from this combined probability using the algorithm described previously except with a high probability cutoff of 0.9 instead of 0.95. The posterior probabilities from the missing data phylo-HMM were also converted into state assignments using a similar method. However, the initial seeds were defined with a cutoff of 0.95, and the seeds were expanded outward to first non-gap symbol or until the posterior probability of the ``missing data'' state was below 0.5. These assignments, which are available in the supplementary data, can be used to filter segments or entire sequences from downstream analyses.
